---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sciCNV)
library(base)
library(robustbase)
library(devtools)
library(Seurat)
library(dplyr)
library(Matrix)
library(qlcMatrix)
library(devtools)
library(svd)
library(ggplot2)
library(ggridges)
library(viridis)
require(scales)
library(RColorBrewer)
library(Rtsne)
library(reticulate)
library(umap)
library(parallel)
library(beanplot)
library(dichromat)
source_url("https://raw.githubusercontent.com/obigriffith/biostar-tutorials/master/Heatmaps/heatmap.3.R")

##
path.code <- "../R"
source(file.path(path.code, "Mito_umi_gn.R"))
source(file.path(path.code, "RTAM_normalization.R"))
source(file.path(path.code, "CNV_infer.R"))
source(file.path(path.code, "Scaling_CNV.R"))
source(file.path(path.code, "CNV_score.R"))
source(file.path(path.code, "sciCNV.R"))
source(file.path(path.code, "heatmap.3.R"))
source(file.path(path.code, "Sketch_AveCNV.R"))
source(file.path(path.code, "CNV_htmp_glist.R"))
source(file.path(path.code, "CNV_htmp_gloc.R"))
source(file.path(path.code, "Opt_MeanSD_RTAM1.R"))
source(file.path(path.code, "Opt_MeanSD_RTAM2.R"))
source(file.path(path.code, "heatmap_break_glist.R"))
source(file.path(path.code, "heatmap_break_gloc.R"))

```



Reading raw data with a list of genes on the first column

```{r}
raw.data1 <- read.table( "../inst/extdata/Sample_20_CPCs__with__20_NPCs.txt", sep = '\t',header = TRUE) 
raw.data2 <- as.matrix(raw.data1[ , -1])
rownames(raw.data2) <- raw.data1[ , 1]
colnames(raw.data2) <- colnames(raw.data1[ , -1])
##
W <- ncol(raw.data2)
No.test <- 20                             # Number of test cells
No.control <- ncol(raw.data2) - No.test   # Number of control cells
Col_Sum <- t(as.numeric(colSums(raw.data2)))
```



####################################################
Quality Control (QC): Elimination of damaged cells 
####################################################

Calculation of total transcript count per cell, idnetified by unique molecular identifiers (UMI)
Calcuation of mitochondrial transcript content per cell

```{r}

nUMI <- t(as.numeric(colSums(raw.data2)))
colnames(nUMI) <- colnames(raw.data2)

mito.genes1 <- read.table( "../inst/extdata/Sample_20_CPCs___Mitochondrial.txt", sep = '\t',header = TRUE)
mito.genes <- as.matrix(mito.genes1[,-1])
percent.mito.G <- t(as.matrix(colSums(mito.genes)))/ ( Col_Sum[1:No.test] + colSums(mito.genes))

nGene1 <- matrix(0, ncol = ncol(raw.data2) , nrow = 1)
nonzero <- function(x){ sum(x != 0) }

nGene1 <- lapply( 1:ncol(raw.data2), function(i){ nonzero(raw.data2[, i])} ) 
nGene1 <- t(as.numeric(nGene1))
colnames(nGene1) <- colnames(raw.data2)
```



---------------------------------
Identification of Damaged Cells
---------------------------------

```{r}
MMS <- CreateSeuratObject(counts = raw.data2, project = "Sample1")
nUMI1 <- as.matrix(nUMI[1:No.test])
nGene2 <- as.matrix(nGene1[1:No.test])
damaged_cells <- Mito_umi_gn( mat=MMS, 
                              percent.mito.G=percent.mito.G,
                              nUMI=nUMI1,
                              nGene=nGene2,
                              No.test=No.test,
                              drop.mads=3)
```



------------------------------
Exclusion of Damaged Cells
------------------------------

```{r}
if( length(damaged_cells) > 0 ){
  Outliers <- damaged_cells
  raw.data <- raw.data2[, - Outliers]
  nUMI <- t(as.matrix(nUMI))[- Outliers]
} else{
  raw.data <- raw.data2
}

nUMI <- as.matrix(nUMI)
raw.data <- as.matrix(raw.data )
rownames(raw.data) <- raw.data1[ , 1]
colnames(nUMI) <- colnames(raw.data)
```

----------------------------------------------------
Sorting of cells by UMI (from largest to smallest)
----------------------------------------------------

```{r}
if(No.control > 0){
raw.data <- raw.data2[, c(colnames(nUMI)[order(as.data.frame(nUMI)[1:No.test], decreasing=TRUE)]
                        ,colnames(nUMI)[No.test+order(as.data.frame(nUMI)[(No.test+1):ncol(raw.data2)],                              decreasing=TRUE)]), drop=FALSE]
} else {
  raw.data <- raw.data2[, colnames(nUMI)[order(as.data.frame(nUMI)[1, ], decreasing=TRUE)], drop=FALSE]
}
raw.data <- as.matrix(raw.data)
rownames(raw.data) <- rownames(raw.data2) 
```

----------------------------
RTAM1/2 Normalziation
----------------------------

```{r}
norm.data <- RTAM_normalization(mat = raw.data,            
                                method = "RTAM2",      
                                Min_nGn  = 250,       
                                Optimizing = FALSE)
rownames(norm.data) <- rownames(raw.data2) 
colnames(norm.data) <- colnames(raw.data)
```


Plotting the normalized expression values of detected genes (y-axis) by cell (X-axis)


```{r}
plot.new()
par(mar=c(5,5,4,2)+1,mgp=c(3,1,0))
par(xaxs="i", yaxs="i") 
par(bty="l")
plot(matrix(1,ncol=1,nrow=nrow(as.matrix(norm.data[,1][norm.data[,1]>0]))), 
     log2(as.matrix(norm.data[,1][norm.data[,1]>0]) +1  ), pch=16, cex=0.3, 
     col="darkgray" ,
     xlim=c(-ncol(norm.data)*.01, ncol(norm.data)*1.01),
     ylim=c(0,  4.2), xlab = "Cells (ranked by UMI)",
     ylab = expression("Normalized gene expression ("*Log[2]*"(.+1 ))"))

for(i in 2:ncol(norm.data)){
  par(new=TRUE)
  points(  matrix(i,ncol=1,nrow=nrow(as.matrix(norm.data[,i][norm.data[,i]>0]))), 
           log2(as.matrix(norm.data[,i][norm.data[,i]>0]) +1  ), pch=16, cex=0.3, 
           col="darkgray")
}
title( paste("Sample1, RTAM2-normalization, cutoff ", 250," nGene ",250,sep=""), 
       col.main = "brown")
```



Plotting the average expression of commonly expressed genes (expressed by >95% cells) for each cell

```{r}
Sqnce <- seq(1,ncol(norm.data),1)
Common.mat <-   as.matrix(norm.data[which(rowSums(norm.data[,Sqnce ] != 0) > 0.95*ncol(norm.data) ), ] )


plot.new()
par(mar=c(5,5,4,2)+1,mgp=c(3,1,0))
par(xaxs="i", yaxs="i") 
par(bty="l")
plot(matrix(1,ncol=1,nrow=nrow(as.matrix(norm.data[,1][norm.data[,1]>0]))), 
     log2(as.matrix(norm.data[,1][norm.data[,1]>0]) +1  ), pch=16, cex=0.3, 
     col="darkgray" ,
     xlim=c(-ncol(norm.data)*.01, ncol(norm.data)*1.01),
     ylim=c(0,  4.2), xlab = "Cells (ranked by UMI)",
     ylab = expression("Normalized gene expression ("*Log[2]*"(.+1 ))"))

for(i in 2:ncol(norm.data)){
  par(new=TRUE)
  points(  matrix(i,ncol=1,nrow=nrow(as.matrix(norm.data[,i][norm.data[,i]>0]))), 
           log2(as.matrix(norm.data[,i][norm.data[,i]>0]) +1  ), pch=16, cex=0.3, 
           col="darkgray")
}
for(j in 1:ncol(norm.data)){
  par( new=TRUE)
  points(matrix(j,ncol=1,nrow=1),
         log2(mean(as.matrix(Common.mat[,j][Common.mat[,j]>0])) + 1 ) ,
         col="red" , pch=15, cex =0.5      
  )
} 
legend(0,0.75,bty="n",pch=16,col=c("red",NA), cex=1.5, legend=paste("Mean of commonly-expressed genes"))
```

plotting the average expression of housekeeping genes

```{r}
Houskeeping_gene_list <- read.table( "../data/HouseKeepingGenes.txt", sep = '\t',header = TRUE)
HK_mat  <- norm.data[which(rownames(norm.data)%in% t(as.matrix(Houskeeping_gene_list))), ]
HK_mat <- as.matrix(HK_mat)
colnames(HK_mat) <- colnames(norm.data)

Mean_HK_mat <- matrix(0, ncol = ncol(norm.data) , nrow = 1)
for(k in 1: (ncol(norm.data))){
  Mean_HK_mat[1,k] <- as.numeric(mean(HK_mat[,k][HK_mat[,k]>0]) )
}


plot.new()
par(mar=c(5,5,4,2)+1,mgp=c(3,1,0))
par(xaxs="i", yaxs="i") 
par(bty="l")
plot(matrix(1,ncol=1,nrow=nrow(as.matrix(norm.data[,1][norm.data[,1]>0]))), 
     log2(as.matrix(norm.data[,1][norm.data[,1]>0]) +1  ), pch=16, cex=0.3, 
     col="darkgray" ,
     xlim=c(-ncol(norm.data)*.01, ncol(norm.data)*1.01),
     ylim=c(0,  4.2), xlab = "Cells (ranked by UMI)",
     ylab = expression("Normalized gene expression ("*Log[2]*"(.+1 ))"))

for(i in 2:ncol(norm.data)){
  par(new=TRUE)
  points(  matrix(i,ncol=1,nrow=nrow(as.matrix(norm.data[,i][norm.data[,i]>0]))), 
           log2(as.matrix(norm.data[,i][norm.data[,i]>0]) +1  ), pch=16, cex=0.3, 
           col="darkgray")
}
par( new=TRUE)
points(log2(Mean_HK_mat[1,] +1 ), col="blue" , pch=15, cex =0.5 )
legend(0,0.5,bty="n",pch=16,col=c("blue",NA), cex=1.5, legend=paste("Mean of Houskeeping genes"))
```

------------------------------------------
Excluding non-expressed genes & clustering
------------------------------------------

```{r}
train1 <- as.matrix(norm.data)
colnames(train1) <- colnames(norm.data)
train <- as.matrix(train1[which(rowSums(train1 != 0) >= 1  ), ] )
```

---------------------------------
Clustering the normalized data
---------------------------------

Finding matrix of single cells (MSC) as a seurat object for clustering

```{r}

MSC <- CreateSeuratObject(counts = train, project = "sample1")
MSC <- FindVariableFeatures(object = MSC)
MSC  <- ScaleData(MSC , verbose = FALSE)
MSC  <- RunPCA(MSC , npcs = 5, verbose = FALSE)
MSC  <- RunUMAP(MSC , reduction = "pca", dims = 1:5)
MSC  <- FindNeighbors(MSC , reduction = "pca", dims = 1:5)
MSC  <- FindClusters(MSC , resolution = 0.5)

MSC <- RunUMAP(MSC, dims = 1:5)
DimPlot(MSC, reduction = "umap", pt.size = 3, label = TRUE, label.size = 4)
```

------------------------------------------------------------------------------------------------------
Running the sciCNV function to derive CNV profiles for single cells from RTAM-normalized scRNA-seq data
------------------------------------------------------------------------------------------------------

```{r}
tst.index  <- seq(1, No.test , 1)                      # No of test cells
ctrl.index <- seq(No.test+1,No.test+No.control , 1) 

genLoc <- read.table( "../data/10XGenomics_gen_pos_GRCh38-1.2.0.txt", sep = '\t', header=TRUE)
genLoc1 <- genLoc 
Spec.genes <- which(as.matrix(genLoc[,1]) %in% as.matrix(rownames(norm.data)))
genLoc <-  genLoc[Spec.genes, c(1,2)]


Ave_NCs <- as.matrix(rowMeans(as.matrix(norm.data[ , ctrl.index ]) ))
```

Generating infered-CNV data for (test and/or control) cells 

```{r}
CNV.data <- sciCNV(norm.mat = norm.data, 
                   ave.ctrl = Ave_NCs,
                   No.test = No.test, 
                   gen.Loc = genLoc,
                   sharpness  = 1,
                   baseline_adj = FALSE, 
                   baseline = 0)
```

------------------------------------------------------------------
Scaling the sciCNV curves and setting a noise filter threshold
------------------------------------------------------------------

Scaling CNV-curves to adjust one copy number gain/losses to height +1/-1 if applicable

```{r}
CNV.data.scaled <- Scaling_CNV(V7Alt = CNV.data, 
                               n.TestCells = No.test, 
                               scaling.factor = 1)
```

In case one needs to re-scale data can change the scaling.factor and run the Scaling.CNV function again

```{r}
CNV.data.scaled <- Scaling_CNV(CNV.data, 
                               n.TestCells = No.test, 
                               scaling.factor = 0.95)
```


Defining M_NF as Noise-Free Matrix of test cells (and control cells)

```{r}
M_NF <- CNV.data.scaled
```

Noise Filteration after scaling

```{r}
noise.thr = 0.2   # Noise threshold
for(w in 1:ncol(M_NF) ){
  for(j in 1:nrow(M_NF)){
    if( (M_NF[j,w] > -noise.thr) && (M_NF[j,w] < noise.thr)  ){
      M_NF[j,w] <- 0
    }
  }
}
M_NF <- as.matrix(M_NF)
```

Taking Square Rate

```{r}
for(w in 1:ncol(M_NF)){
  for(j in 1:nrow(M_NF)){
    if (M_NF[j,w] > 0){
      M_NF[j,w]<- sqrt( as.numeric(M_NF[j,w]))
    } else if (M_NF[j,w] < 0){
      M_NF[j,w]<- -sqrt( -as.numeric(M_NF[j,w]))
    }
  }
}

rownames(M_NF) <- rownames(CNV.data.scaled)
colnames(M_NF) <- c(colnames(CNV.data.scaled)[-length(colnames(CNV.data.scaled))], "AveTest")
```

Assigning chromosome number to each gene sorted based on chromosome number, starts and ends to sketch the average sciCNV curve of test cells

```{r}
Specific_genes <- which( genLoc[, 1] %in% rownames(CNV.data.scaled))
Assoc.Chr <-  genLoc[Specific_genes, 2]
#Assoc.Chr <-  apply(Assoc.Chr, 2, as.numeric)
```

Finalizing the sciCNV-matrix by attaching gene-name and chromosome number lists

```{r}
M_NF1 <- cbind(as.matrix(genLoc[Specific_genes, 1]), 
               as.matrix(genLoc[Specific_genes, 2]), 
               M_NF)
colnames(M_NF1) <- c("Genes", "Chromosomes", colnames(norm.data), "Ave test")
M_NF2 <- as.matrix(M_NF1)
M_NF3 <- as.matrix(M_NF2[ , ncol(M_NF2)] )
rownames(M_NF3) <- as.matrix(M_NF2[ , 1] )
```

Sketching the final sciCNV result

```{r}
Sketch_AveCNV( Ave.mat = M_NF[, ncol(M_NF)], Assoc.Chr=Assoc.Chr )
```


------------------------------------------------------------------  
Calculating tumor CNV scores to segregate normal vs tumor cells
------------------------------------------------------------------

Providing the umor-score 

```{r}
TotScore <- CNV_score( M_nf = M_NF )
```

Sketching tumor scores for all cells showing segregation of test/control tumor scores

```{r}
TotScoreSort0 <- sort(TotScore[1,1:No.test])
TotScoreSortNPC <- sort(TotScore[1,(No.test+1):ncol(TotScore)])
Labels <- c("Control","Test")
names <- as.matrix(c(rep(Labels[1], length(TotScoreSortNPC)), rep(Labels[2],length(TotScoreSort0) )) )
value <- c(as.matrix(TotScoreSortNPC), as.matrix(TotScoreSort0))
data=data.frame(names,value)
data$factor <- factor(data$names, levels=Labels)
##
graphics.off()
plot.new()
par(mar=c(5,5,4,2)+1,mgp=c(3,1,0))
par(bty="l")
COL <- c( "lightgrey","royalblue1","royalblue1","royalblue1")
beanplot::beanplot(data$value ~ data$factor , col=alpha(COL,0.8),
         ylim=c(min(0,min(TotScore)*1.2),max(TotScore)*1.2),
         xlab = "Cell type",
         ylab = "CNV score",
         what=c(0,1,1,1),
         bty='l', boxcol="gray" ,
         outpch=16, outcex=1)
title("Sample1: CNV score of individuals", col.main = "brown")
```
 
--------------------------------------------------------------
Heatmap of sciCNV profiles and detection of subclones
--------------------------------------------------------------
generating heatmap:
In order to use genomic locations in our heatmap we read Gen.Loc matrix 
with list of genes, chromosome numbers, starts and ends:

```{r}
CNV.matrix <- t( M_NF[, -ncol(M_NF)])   
rownames(CNV.matrix) <- colnames(M_NF[, -ncol(M_NF)])
colnames(CNV.matrix) <- rownames(CNV.data)

## ## Heatmap of sciCNV plotted against list of genes
breakGlist <- rep(0, 24)
breakGlist <- heatmap_break_glist(CNV.mat2 = CNV.matrix )

##
# CNV_htmp_glist( CNV.mat2 = CNV.matrix,
#                clustering = FALSE,
#                sorting = TRUE,
#                CNVscore = TotScore,
#                cluster.lines = NULL,
#                breakGlist = breakGlist,
#                No.test = No.test
#                )
```

Heatmap of sciCNV plotted against genomic location

```{r}
breakGloc <- rep(0, 24)
breakGloc <- heatmap_break_gloc(CNV.mat2 = CNV.matrix)
#
# CNV_htmp_gloc( CNV.mat2 = CNV.matrix,
#                clustering = FALSE,
#                sorting = TRUE,
#                CNVscore = TotScore,
#                cluster.lines = NULL,
#                breakGloc = breakGloc,
#                No.test = No.test)

```

